---
title: "My title"
subtitle: "My subtitle if needed"
author: 
  - Mandy He
thanks: "Code and data are available at: https://github.com/MandyHe7/Toronto-Bids-Awarded-Contracts"
date: November 27 2024
date-format: long
abstract: "This study investigates supplier equity within Toronto’s public procurement system, focusing on the representation of small businesses and historically underrepresented sectors. Leveraging data from the Toronto Bids Awarded Contracts dataset, the analysis examines award distributions, contract values, and supplier diversity. The study aims to identify potential disparities in supplier representation and evaluate the extent to which public contracts promote equitable opportunities. Recommendations are provided to enhance supplier inclusion and foster a fairer procurement environment, emphasizing transparency, outreach, and support for underrepresented businesses."
format: pdf
number-sections: true
bibliography: references.bib
toc: true 
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(dplyr)
library(ggplot2)
library(scales)
library(lubridate)
library(knitr)
library(kableExtra)
library(broom)


# load data
Contract_data <- read_csv(here::here("data/02-analysis_data/analysis_data.csv"))

```


# Introduction

Equitable public procurement is a cornerstone of inclusive economic development. The City of Toronto, through its competitive contracting process, allocates significant resources to suppliers for goods and services. These contracts serve as vital economic lifelines, particularly for small businesses and sectors that have traditionally been underrepresented in public procurement. However, ensuring equitable access to these opportunities remains a challenge, as systemic barriers may limit the participation of diverse suppliers.

This study explores the equity of supplier representation in Toronto’s public contracts, focusing on small businesses and underrepresented sectors. Using the Toronto Bids Awarded Contracts dataset, the research seeks to answer critical questions: Are small businesses and diverse suppliers equitably represented in contract awards? What trends emerge in terms of award amounts and supplier participation? By addressing these questions, the study aims to provide actionable insights to promote fair and inclusive procurement practices.

The research contributes to a growing discourse on supplier diversity and public accountability, underscoring the importance of fostering inclusive economic opportunities in municipal contracting. This is particularly relevant in a diverse city like Toronto, where equitable procurement practices can significantly impact local communities and drive sustainable development. Through this investigation, the study highlights current gaps and offers recommendations to strengthen supplier equity in Toronto’s public procurement system.

The remainder of this paper is structured as follows. @sec-data will discuss the data used in the study, highlighting key aspects and limitations. @sec-Results will focus on generating graphs using the data presented in @sec-data. @sec-discussion will elaborate on the graphs from @sec-Results, providing interpretations and possible explanations for the findings. 


# Data {#sec-data}

## Overview

The data used in this paper is access from Open Data Toronto and the particular data set used was the Toronto Bids Awarded Contracts [@toronto_tobids]. To analysis the data and creating graphs using the data, following package that was build in the statistical programming language R [@citeR] was used: tidyverse [@tidyverse], dplyr [@dplyr], lubridate [@lubridate], scales [@scales], knitr [@knitr], kableExtra [@kableExtra], broom[@broom] and ggplot2 [@ggplot2]. LLM [@openai2023chatgpt] was also use in term of helping with coding in R to graph data, simulating data, cleaning data, and revising writing. 

@tbl-Table1 is a preview of the clean data that will be used through out the analysis of this paper. 

```{r}
#| label: tbl-Table1
#| tbl-cap: Data Preview
#| echo: false
#| warning: false
#| message: false


# Display the table preview
kable(head(Contract_data), 
      caption = "Data Preview", 
      align = 'l', 
      col.names = c("Unique ID", "Document Number", "RFx Type", 
                    "High Level Category", "Successful Supplier", 
                    "Awarded Amount ($)", "Award Date", 
                    "Division", "Small Business"))  %>%
  kable_styling(font_size = 10) # Adjust font size 

```

## Measurement
	
To understand how a real-world phenomenon transitions into a structured dataset, it is essential to examine the lifecycle of data collection, organization, and analysis. In the case of the Toronto Bids Awarded Contracts dataset, the phenomenon begins with the City's procurement process—a critical operation where goods, services, and construction are solicited and acquired to meet public needs. This process involves multiple stages, from identifying the need for procurement, issuing requests for proposals or tenders, evaluating bids, and awarding contracts to successful suppliers. Each of these steps reflects a real-world activity that, when documented, forms the foundation for the dataset.

The transition starts with the City identifying a need, such as infrastructure development, service improvement, or administrative support. This leads to the creation of a solicitation document, outlining the scope of work, evaluation criteria, and requirements. Suppliers submit their bids, and the City evaluates them based on predefined criteria, such as cost, quality, and adherence to specifications. Once a contract is awarded, key details—including the awarded amount, supplier name, and the division requesting the procurement—are recorded.

This information is entered into the City's procurement management system, capturing the structured attributes necessary for documentation and analysis. For instance, details like the solicitation type (e.g., RFP, RFQ), high-level category (e.g., Construction Services, Professional Services), and award date are meticulously logged. To enhance the dataset's utility, additional attributes such as the division involved, buyer contact information, and a summary of the solicitation are included. The creation of a dummy variable, such as the classification of small businesses, involves further data processing to derive new insights from raw attributes like the awarded amount.

Ultimately, the dataset represents a structured snapshot of a complex real-world process. Each entry reflects decisions, negotiations, and activities that unfolded in the real world, carefully abstracted into numerical and categorical variables. This process of translation allows researchers, policymakers, and the public to analyze the data to uncover patterns, monitor equity, and assess the efficiency and fairness of the City's procurement system. The dataset, while detailed, is always an abstraction—capturing key facets of the phenomenon while omitting others deemed less relevant to the analysis.

## Data Variables

The dataset includes information on competitive procurement processes and awarded contracts by the City of Toronto. Key variables are:

- Unique Row Identifier (_id): A unique identifier for each record in the Open Data database.

- Composite Key (unique_id): A unique identifier combining related procurement details.

- Document Number: A reference number used by the City to track the solicitation during the procurement process.

- RFx Type (Solicitation Type): The procurement method used for the solicitation, such as Request for Proposal (RFP), Request for Quotation (RFQ), Tender (RFT), and others. This indicates the process used to seek bids or proposals from suppliers.

- High-Level Category: Groups the goods or services being procured, such as Goods and Services, Professional Services, or Construction Services.

- Successful Supplier: The name of the supplier awarded the contract.

- Awarded Amount: The total monetary value awarded to the supplier for the contract.

- Award Date: The date the contract between the City and the supplier was finalized.

- Division: The City Division requesting the procurement. The awarded supplier will have a contract with this division.

- Buyer Details: Includes the buyer's name, email, and phone number as the primary purchasing contact for the solicitation or contract.

- Solicitation Document Description: A summary of the solicitation, providing additional context about the procurement request.

In addition, there was a small business classification was created for its role in the analysis.

- Small Business (Dummy Variable): A binary variable indicating whether a supplier qualifies as a small business. Businesses are classified as small (TRUE), while others are classified as large (FALSE) determine by ChatGPT analysis, refer to [Appendix -@sec-appdata] for more detail.


# Results

The pie chart @fig-PieChart demonstrates majority of contracts awarded to small businesses (marked as TRUE for small business). This indicates that small businesses have relative representation in the awarding process, suggesting equity just by looking at proportion.
```{r}
#| label: fig-PieChart
#| fig-cap: Proportion of Small Business Awards
#| echo: false
#| warning: false
#| message: false

# Pie chart: Proportion of Small Business Awards
small_business_summary <- Contract_data %>%
  group_by(Small_Business) %>%
  summarise(Total_Awarded = sum(Awarded_Amount, na.rm = TRUE))

ggplot(small_business_summary, aes(x = "", y = Total_Awarded, fill = Small_Business)) +
  geom_bar(stat = "identity", width = 1, color = "white") +
  coord_polar(theta = "y") +
  labs(title = "Proportion of Small Business Awards",
       x = NULL,
       y = NULL) +
  theme_void()

```

\newpage
This bar chart @fig-bargraph displays the total awarded amount ($) for different RFx (Request for X) types. Key observations:

RFT (Request for Tender) has the highest awarded amount, exceeding $1.2 billion, suggesting that most of the procurement budget is funneled through tenders.
RFQ (Request for Quotation) and RFP (Request for Proposal) follow, with awarded amounts slightly below $500 million and around $400 million, respectively.
Other RFx types (e.g., NRFP, OTP, RFSQ) have minimal awarded amounts, indicating limited use in contract awards.
```{r}
#| label: fig-bargraph 
#| fig-cap: Total Awarded Amount by RFx Type
#| echo: false
#| warning: false
#| message: false

# Summarize awarded amounts by RFx Type
rfx_summary <- Contract_data %>%
  group_by(RFx_Type) %>%
  summarise(Total_Awarded = sum(Awarded_Amount, na.rm = TRUE)) %>%
  arrange(desc(Total_Awarded))

# Bar plot: Total Awarded Amount by RFx Type
ggplot(rfx_summary, aes(x = RFx_Type, y = Total_Awarded, fill = RFx_Type)) +
  geom_bar(stat = "identity", show.legend = FALSE) +
  scale_y_continuous(labels = dollar_format()) +
  labs(title = "Total Awarded Amount by RFx Type",
       x = "RFx Type",
       y = "Total Awarded Amount ($)") +
  theme_minimal()
```


\newpage
This stacked bar chart @@fig-bar2 breaks down supplier classification (Small Business vs. Large Business) for each RFx type:

Small Businesses (TRUE) dominate RFQ, RFP, and RFSQ types, showing significant representation in these categories.
Large Businesses (FALSE) appear to win the majority of contracts in RFT, reflecting a potential preference for larger suppliers in high-value tenders.
NRFP and OTP show exclusive or near-exclusive awards to Large Businesses, albeit with lower overall amounts.

```{r}
#| label: fig-bar2
#| fig-cap: Proportion of Small vs. Large Businesses by RFx Type
#| echo: false
#| warning: false
#| message: false

# Calculate proportions of Small and Large Businesses by RFx Type
rfx_small_large <- Contract_data %>%
  group_by(RFx_Type, Small_Business) %>%
  summarise(Total_Awarded = sum(Awarded_Amount, na.rm = TRUE)) %>%
  mutate(Proportion = Total_Awarded / sum(Total_Awarded)) %>%
  ungroup()

# Bar Plot: Proportion of Small vs. Large Businesses by RFx Type
ggplot(rfx_small_large, aes(x = RFx_Type, y = Proportion, fill = Small_Business)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(title = "Proportion of Small vs. Large Businesses by RFx Type",
       x = "RFx Type",
       y = "Proportion",
       fill = "Small Business") +
  theme_minimal()

```

This bar graph @fig-bar3 displays the total awarded amounts split between small (TRUE) and large (FALSE) businesses across different RFx types. The results show a stark contrast in how contract awards are distributed. For RFP (Request for Proposal) and RFQ (Request for Quotation), small businesses receive a comparable share of awarded amounts, indicating these procurement types may be more accessible to smaller enterprises. However, for RFT (Request for Tender), which accounts for the largest total awarded amount, the majority of awards go to large businesses. In contrast, RFx types such as NRFP (Non-Request for Proposal) and RFSQ (Request for Supplier Qualification) have minimal total awarded amounts and show varying levels of small business participation.


```{r}
#| label: fig-bar3
#| fig-cap: Total Awarded Amount by Small vs. Large Businesses for each RFx Type
#| echo: false
#| warning: false
#| message: false

# Calculate total awarded amounts by RFx Type and Small Business status
rfx_award_distribution <- Contract_data %>%
  group_by(RFx_Type, Small_Business) %>%
  summarise(Total_Awarded = sum(Awarded_Amount, na.rm = TRUE)) %>%
  ungroup()

# Bar Plot: Total Awarded Amount by Small vs. Large Businesses for each RFx Type
ggplot(rfx_award_distribution, aes(x = RFx_Type, y = Total_Awarded, fill = Small_Business)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_y_continuous(labels = scales::dollar_format()) +
  labs(title = "Total Awarded Amount by Small vs. Large Businesses for Each RFx Type",
       x = "RFx Type",
       y = "Total Awarded Amount ($)",
       fill = "Small Business") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```



This box plot @fig-Boxplot shows the distribution of awarded contract amounts across various high-level procurement categories, such as Construction Services, Goods & Services, Offer to Purchase, and Professional Services.
Observations:
Construction Services has the widest range of contract amounts, with several high-value outliers above $75 million.
Other categories, like Professional Services and Goods & Services, have much smaller awarded amounts overall, with tighter distributions and fewer high-value outliers.
Offer to Purchase appears to have minimal representation in terms of awarded amounts.

```{r}
#| label: fig-Boxplot
#| fig-cap: Awarded Amounts by High-Level Category
#| echo: false
#| warning: false
#| message: false

# Box plot: Awarded Amounts by High-Level Category
ggplot(Contract_data, aes(x = High_Level_Category, y = Awarded_Amount, fill = High_Level_Category)) +
  geom_boxplot(show.legend = FALSE) +
  scale_y_continuous(labels = dollar_format()) +
  labs(title = "Distribution of Awarded Amounts by High-Level Category",
       x = "Category",
       y = "Awarded Amount ($)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```

This bar chart @fig-bar4 visualizes the proportion of contracts awarded to small businesses (in blue) versus large businesses (in red) across the same high-level procurement categories.
Observations:
Construction Services and Professional Services have a significant portion of contracts awarded to large businesses, indicating potential barriers for small business participation.
In contrast, Goods & Services and Offer to Purchase categories have a higher representation of small businesses, suggesting that these categories are more accessible to smaller enterprises.
```{r}
#| label: fig-bar4
#| fig-cap: Proportion of Small vs. Large Businesses by High-Level Category
#| echo: false
#| warning: false
#| message: false


# Calculate proportions of Small and Large Businesses by High-Level Category
category_small_large <- Contract_data %>%
  group_by(High_Level_Category, Small_Business) %>%
  summarise(Total_Awarded = sum(Awarded_Amount, na.rm = TRUE)) %>%
  mutate(Proportion = Total_Awarded / sum(Total_Awarded)) %>%
  ungroup()

# Bar Plot: Proportion of Small vs. Large Businesses by High-Level Category
ggplot(category_small_large, aes(x = High_Level_Category, y = Proportion, fill = Small_Business)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(title = "Proportion of Small vs. Large Businesses by High-Level Category",
       x = "High-Level Category",
       y = "Proportion",
       fill = "Small Business") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

This graph @fig-bar6 underscores the substantial role small businesses play in construction and professional services, challenging assumptions that large firms consistently dominate high-value contracts. However, the absence of awarded amounts under "Offer to Purchase" suggests the need for further investigation into how this category functions within procurement practices.

```{r}
#| label: fig-bar6
#| fig-cap: Total Awarded Amount by Small vs. Large Businesses for each RFx Type
#| echo: false
#| warning: false
#| message: false

# Calculate total awarded amounts by RFx Type and Small Business status
rfx_award_distribution <- Contract_data %>%
  group_by(High_Level_Category, Small_Business) %>%
  summarise(Total_Awarded = sum(Awarded_Amount, na.rm = TRUE)) %>%
  ungroup()

# Bar Plot: Total Awarded Amount by Small vs. Large Businesses for each RFx Type
ggplot(rfx_award_distribution, aes(x = High_Level_Category, y = Total_Awarded, fill = Small_Business)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_y_continuous(labels = scales::dollar_format()) +
  labs(title = "Total Awarded Amount by Small vs. Large Businesses for Each High Level Category",
       x = "High Level Category",
       y = "Total Awarded Amount ($)",
       fill = "Small Business") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

This line graph @fig-Timeseries illustrates the awarded amounts to small (labeled "TRUE") and large (labeled "FALSE") businesses over time (on a monthly basis). Key observations include:

Dominance of Large Businesses: Large businesses consistently secure higher awarded amounts than small businesses throughout the analyzed time period.
Fluctuations in Awarded Amounts: Both business types experience fluctuations in awarded amounts over time, with notable peaks for large businesses, particularly in mid-2023 and mid-2024.
Growth for Small Businesses: Despite lower totals overall, the awarded amounts for small businesses show occasional spikes, indicating their competitive presence in certain months.
This graph highlights the disparity in procurement awards, suggesting that while small businesses participate, they often face structural disadvantages compared to larger entities.
```{r}
#| label: fig-Timeseries
#| fig-cap: Awarded Amounts Over Time
#| echo: false
#| warning: false
#| message: false

# Load necessary libraries
library(ggplot2)
library(dplyr)
library(lubridate)


# Summarize awarded amounts by month and business type
time_series_data <- Contract_data %>%
  mutate(Month = floor_date(Award_Date, "month")) %>%
  group_by(Month, Small_Business) %>%
  summarise(Total_Awarded = sum(Awarded_Amount, na.rm = TRUE)) %>%
  ungroup()

# Plot the time series
ggplot(time_series_data, aes(x = Month, y = Total_Awarded, color = Small_Business)) +
  geom_line(size = 1.2) +
  labs(title = "Awarded Amount to Small vs. Large Businesses Over Time",
       x = "Time (Monthly)",
       y = "Total Awarded Amount ($)",
       color = "Business Type") +
  theme_minimal() +
  scale_y_continuous(labels = scales::dollar_format()) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

This scatter plot with a regression line @fig-TimeseriesSmall focuses on the awarded amounts to small businesses over time and reveals:

Positive Trend: The red regression line shows a gradual increase in awarded amounts to small businesses over time, suggesting a slight improvement in their procurement outcomes.
Variance in Data: The wide spread of data points around the regression line indicates variability in the awarded amounts, implying that certain months see much higher or lower awards than others.
This graph suggests an encouraging trend for small businesses but also highlights inconsistency, underscoring the need for policy interventions to ensure sustained growth.
```{r}

#| label: fig-TimeseriesSmall
#| fig-cap: Awarded Amounts Over Time
#| echo: false
#| warning: false
#| message: false

small_business_data <- Contract_data %>% 
  filter(Small_Business == TRUE) %>% 
  mutate(Month = floor_date(Award_Date, "month"))

# Aggregate awarded amounts by month for small businesses
regression_data <- small_business_data %>%
  group_by(Month) %>%
  summarise(Total_Awarded = sum(Awarded_Amount, na.rm = TRUE)) %>%
  ungroup()

# Add a numeric time variable for regression
regression_data$Time <- as.numeric(regression_data$Month)

# Fit a linear regression model
model <- lm(Total_Awarded ~ Time, data = regression_data)
summary(model)

# Plot the results
ggplot(regression_data, aes(x = Month, y = Total_Awarded)) +
  geom_point(color = "blue", size = 2) +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  labs(title = "Regression Analysis: Awarded Amount to Small Businesses Over Time",
       x = "Time (Monthly)",
       y = "Total Awarded Amount ($)") +
  theme_minimal() +
  scale_y_continuous(labels = scales::dollar_format()) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```






A logistic regression model can help assess the factors influencing whether a contract is awarded to a small business.

In regression result @tbl-Table2, examines how factors like RFx type, high-level category, and awarded amount affect small business contract awards. Key findings include:

Significant Barriers: Larger awarded amounts and categories like "Goods and Services" and "Professional Services" negatively impact small business participation, indicating structural challenges.
RFx Types: No RFx type shows a statistically significant impact, suggesting that other systemic factors (e.g., eligibility criteria) influence outcomes.
Implications: The results highlight barriers for small businesses in high-value contracts, emphasizing the need for targeted reforms to improve access and equity in procurement.

```{r}
#| label: tbl-Table2
#| tbl-cap: Logistic regression model
#| echo: false
#| warning: false
#| message: false

# Logistic regression model
log_model <- glm(Small_Business ~ RFx_Type + High_Level_Category + Awarded_Amount, 
                 data = Contract_data, family = binomial)

# Tidy the model output for a table
log_model_summary <- broom::tidy(log_model)

# Add odds ratios (exponentiated coefficients) and confidence intervals
log_model_summary <- log_model_summary %>%
  mutate(Odds_Ratio = exp(estimate)) %>%
  mutate(Lower_CI = exp(estimate - 1.96 * std.error),
         Upper_CI = exp(estimate + 1.96 * std.error))

# Display the table
log_model_summary %>%
  select(term, estimate, std.error, Odds_Ratio, Lower_CI, Upper_CI, p.value) %>%
  kable(
    caption = "Logistic Regression Results: Predicting Small Business Contracts",
    col.names = c("Term", "Estimate (Log Odds)", "Std. Error", 
                  "Odds Ratio", "Lower CI", "Upper CI", "P-Value"),
    digits = 4, align = "c"
  ) %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"))


```


# Discussion

## Equitable Representation Across RFx Types {#sec-first-point}

The analysis of supplier representation across different RFx types, as shown in the graphs, highlights significant trends and disparities in Toronto's public procurement system. Small businesses are notably well-represented in RFQ (Request for Quotation), RFP (Request for Proposal), and RFSQ (Request for Supplier Qualification) processes, indicating that these procurement methods may offer more accessible entry points due to lower barriers or more flexible requirements. This is reflected in the proportional graph, where small businesses consistently account for a significant share of awarded contracts in these RFx types.

However, the dominance of large businesses in RFT (Request for Tender) contracts reveals an important equity concern. As illustrated in the total awarded amounts graph, RFT contracts account for the highest share of procurement spending, exceeding $1.2 billion. The majority of these funds are awarded to large businesses, signaling structural barriers for smaller firms. RFT contracts often involve large-scale projects or stringent qualification criteria, which can exclude smaller suppliers who lack the resources or capacity to meet these demands. This limited participation by small businesses in RFTs highlights a critical disparity in the distribution of high-value procurement opportunities.

The implications of these findings underscore the urgent need for targeted policy interventions. Simplifying bidding requirements for RFT contracts could lower barriers for small businesses, while creating set-aside opportunities specifically for smaller firms would ensure a more equitable distribution of high-value contracts. Furthermore, policies encouraging larger firms to subcontract with small enterprises could provide an additional pathway for inclusion. Enhanced outreach and support programs—focused on training, guidance, and awareness—would also equip small businesses to compete more effectively in these tenders.

Together, these measures can transform Toronto’s procurement system to better embody the principles of equity and inclusivity. By addressing disparities across RFx types, the city can foster fair opportunities for businesses of all sizes and strengthen the diversity and resilience of its supplier base.


## Supplier Equity Across High-Level Categories {#sec-second-point}

The analysis of awarded contract amounts and supplier participation across high-level procurement categories reveals significant trends relevant to equitable representation in Toronto’s public procurement system.

Disparities in Award Distribution
@fig-Boxplot, illustrating the distribution of awarded amounts by high-level category, shows that Construction Services dominates procurement in terms of contract value. This category includes numerous high-value outliers exceeding $75 million, reflecting the capital-intensive nature of infrastructure projects. These projects often require substantial financial and operational capacity, favoring larger businesses. In contrast, categories like Goods & Services and Professional Services exhibit tighter distributions with lower average awarded amounts, indicating more equitable contract allocations across businesses of various sizes.

@fig-bar6, which examines total awarded amounts for small and large businesses across high-level categories, highlights the dominance of small businesses in Construction Services, with total awards nearing $1.2 billion, compared to $300 million for large businesses. This suggests that small businesses play a significant role and are highly competitive within this high-value category. Similarly, in Goods & Services, small businesses receive slightly less than $300 million, while large businesses account for less than $150 million, again showing strong representation by small enterprises. The Professional Services category mirrors this trend, with small businesses securing more total awarded amounts than large businesses, which is noteworthy given the specialized nature of this category. The Offer to Purchase category, however, has no awarded amounts, suggesting either limited use within the dataset or an underrepresentation in procurement, which warrants further investigation.

Small vs. Large Business Representation
 @fig-bar4, showing proportions of small and large businesses by category, reveals that small businesses maintain strong representation in Construction Services, despite its capital-intensive nature. However, in Professional Services, large businesses show a higher proportion of awards, likely due to specialized technical expertise or operational scale. Goods & Services demonstrates a more balanced representation, where small businesses capture a significant share of contracts, reflecting fewer barriers to entry in this category. The lack of activity in the Offer to Purchase category requires further exploration to understand its relevance within procurement.

Implications for Equity in Procurement
These trends highlight important equity considerations. The strong performance of small businesses in Construction Services challenges assumptions that high-value categories exclusively benefit large firms. Small businesses also exhibit competitive success in Professional Services, disputing the notion that specialized or high-value categories are only accessible to larger companies. Conversely, Goods & Services appears to offer more equitable access to small businesses, further suggesting that lower-value contracts may have fewer entry barriers. However, the Offer to Purchase category, with no awarded amounts, requires deeper examination to determine whether it represents a missed opportunity for small businesses or simply has limited procurement activity.

Recommendations to Address Disparities
To foster greater equity in procurement across high-level categories:

- Simplify Bidding Requirements: Streamlining requirements for Construction Services and Professional Services could lower entry barriers for small businesses, facilitating more inclusive participation.
- Introduce Set-Asides: Allocating a portion of contracts in high-value categories for small or diverse suppliers could improve inclusivity.
- Encourage Subcontracting: Requiring large firms to subcontract work to small businesses distributes economic benefits more broadly.
- Capacity-Building Programs: Offering technical training, financial assistance, and mentorship for small businesses can empower them to compete in high-value categories.


By addressing these challenges, Toronto’s procurement system can better balance representation and ensure fair opportunities for businesses of all sizes and backgrounds. These findings reinforce the importance of refining procurement policies to advance equity and inclusivity across all high-level categories.

## Trends in Small Business and Diverse Supplier Representation {#sec-thrid-point}

The analysis of procurement data sheds light on trends in award amounts and supplier participation, revealing critical insights into the equitable representation of small businesses and diverse suppliers in contract awards.

Unequal Representation in Award Amounts
The "Total Awarded Amount by Small vs. Large Businesses for Each RFx Type" graph reveals a significant disparity in awarded amounts across procurement types. While small businesses show meaningful participation in RFQ (Request for Quotation) and RFP (Request for Proposal) contracts, large businesses dominate high-value RFT (Request for Tender) contracts. This trend suggests that procurement processes for high-value contracts pose barriers for small businesses, possibly due to complex requirements, higher capital needs, or economies of scale favoring larger firms.

Fluctuating Participation Over Time
The "Awarded Amount to Small vs. Large Businesses Over Time" graph indicates that small business participation fluctuates significantly, with occasional spikes in awarded amounts. These spikes suggest temporary efforts to engage small businesses in the procurement process, but the lack of consistency indicates systemic challenges in sustaining equitable participation. Large businesses, in contrast, demonstrate a steadier pattern of dominance across time.

Insights from Regression Analysis
The regression analysis confirms structural inequities. Contracts with higher awarded amounts are less likely to be awarded to small businesses, and certain categories, such as "Professional Services," show a similar bias. While the RFx type itself does not significantly impact the likelihood of small business success, the observed trends point to persistent structural barriers limiting equitable access to high-value opportunities for small and diverse suppliers.

Implications and Recommendations
These findings highlight a need for targeted reforms in Toronto’s procurement practices. Simplifying bidding requirements, particularly for RFT contracts, and introducing small business set-asides or quotas could address systemic inequities. Further, fostering subcontracting relationships between large and small firms and providing support programs for small businesses can enhance their competitiveness. A more equitable representation of small and diverse suppliers not only aligns with principles of inclusivity but also enriches the procurement ecosystem by encouraging innovation and broader economic participation.


## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

\appendix

# Appendix {-}


# Details on Selection of Big Business Out From Data {#sec-appdata}

The reult of selection of big business is by using ChatGPT.

I asked: 
in the data, under the Successful_Supplier, can you anaysis and pull out the excat names that are consider as big business and put them into a big_business_list, so i can use it for r code
ChatGPT said:
ChatGPT
The dataset contains the following columns:

unique_id: A unique identifier for each record.
Document_Number: The associated document number.
RFx_Type: Type of request (e.g., RFQ, RFP).
High_Level_Category: The category of the procurement.
Successful_Supplier: The name of the supplier who won the contract.
Awarded_Amount: The monetary value of the awarded contract.
Award_Date: The date the contract was awarded.
Division: The division responsible for the contract.
Small_Business: Indicates if the supplier is classified as a small business (True/False).
To identify big businesses, I will filter out suppliers where Small_Business is False and create a list of their names.

The following are the big businesses identified from the dataset:

Guillevin International Co.
Sysco
Sysco Toronto, A Division Of Sysco Canada Inc.
Stantec Consulting Ltd
KPMG LLP
KPMG
Bennett Mechanical Installations
Fer-Pal Construction Ltd.
WSP Canada Inc
Black And McDonald Limited
Thermo Fisher Scientific
Sutherland-Schultz Ltd.
Morrison Hershfield Ltd
Morrison Hershfield
Metro Freightliner Hamilton Inc
Ricoh Canada Inc
Altus Group Limited
Kroll Consulting Canada Co.
Suncorp Valuations
Logixx Security Inc
Garda Canada Security Corporation
Drake International Inc
Bevertec CST Inc
AECOM Canada Ltd
Stericycle Ulc
Schindler Elevator Corporation
Graham Bros. Construction Limited
Morrison Hershfield Limited
Softchoice LP
Damen Shipbuilding 5 B.V.
Graham Bros Construction Ltd.
Gartner Canada
Parsons Inc
Jacobs Consultancy Canada Inc
Stericycle, Ulc

This list is now ready for use in your R code. Let me know if you need it in a specific format! 


\newpage


# References


