---
title: "My title"
subtitle: "My subtitle if needed"
author: 
  - Mandy He
thanks: "Code and data are available at: https://github.com/MandyHe7/Toronto-Bids-Awarded-Contracts"
date: November 27 2024
date-format: long
abstract: "This study investigates supplier equity within Toronto’s public procurement system, focusing on the representation of small businesses and historically underrepresented sectors. Leveraging data from the Toronto Bids Awarded Contracts dataset, the analysis examines award distributions, contract values, and supplier diversity. The study aims to identify potential disparities in supplier representation and evaluate the extent to which public contracts promote equitable opportunities. Recommendations are provided to enhance supplier inclusion and foster a fairer procurement environment, emphasizing transparency, outreach, and support for underrepresented businesses."
format: pdf
number-sections: true
bibliography: references.bib
toc: true 
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(dplyr)
library(ggplot2)
library(scales)

# load data
Contract_data <- read_csv(here::here("data/02-analysis_data/analysis_data.csv"))

```


# Introduction

Equitable public procurement is a cornerstone of inclusive economic development. The City of Toronto, through its competitive contracting process, allocates significant resources to suppliers for goods and services. These contracts serve as vital economic lifelines, particularly for small businesses and sectors that have traditionally been underrepresented in public procurement. However, ensuring equitable access to these opportunities remains a challenge, as systemic barriers may limit the participation of diverse suppliers.

This study explores the equity of supplier representation in Toronto’s public contracts, focusing on small businesses and underrepresented sectors. Using the Toronto Bids Awarded Contracts dataset, the research seeks to answer critical questions: Are small businesses and diverse suppliers equitably represented in contract awards? What trends emerge in terms of award amounts and supplier participation? By addressing these questions, the study aims to provide actionable insights to promote fair and inclusive procurement practices.

The research contributes to a growing discourse on supplier diversity and public accountability, underscoring the importance of fostering inclusive economic opportunities in municipal contracting. This is particularly relevant in a diverse city like Toronto, where equitable procurement practices can significantly impact local communities and drive sustainable development. Through this investigation, the study highlights current gaps and offers recommendations to strengthen supplier equity in Toronto’s public procurement system.

The remainder of this paper is structured as follows. @sec-data will discuss the data used in the study, highlighting key aspects and limitations. @sec-Results will focus on generating graphs using the data presented in @sec-data. @sec-discussion will elaborate on the graphs from @sec-Results, providing interpretations and possible explanations for the findings. 


# Data {#sec-data}

## Overview

We use the statistical programming language R [@citeR].... Our data [@shelter].... Following @tellingstories, we consider...

Overview text

```{r}
#| label: tbl-Table1
#| tbl-cap: summary of data
#| echo: false


# Display the table without the 'id' column
kable(Contract_data,
  col.names = c(
    "Year", "Total Decedents",
    "Avg Age of Death (All)", "Male Decedents",
    "Avg Age of Death (Male)", "Female Decedents",
    "Avg Age of Death (Female)", "Trans/NB/2S Decedents",
    "Avg Age of Death (Trans/NB/2S)"
  )
)
```

## Measurement
	
Some paragraphs about how we go from a phenomena in the world to an entry in the dataset.

## Outcome variables

Add graphs, tables and text. Use sub-sub-headings for each outcome variable or update the subheading to be singular.


Talk way more about it. 

## Predictor variables

Add graphs, tables and text.

Use sub-sub-headings for each outcome variable and feel free to combine a few into one if they go together naturally.



# Model

The goal of our modelling strategy is twofold. Firstly,...

Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].

## Model set-up

Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.  

\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}

We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.


### Model justification

We expect a positive relationship between the size of the wings and time spent aloft. In particular...

We can use maths by including latex between dollar signs, for instance $\theta$.



# Results

@fig-bargraph we can see that the graph shows that the biggest amount is awared to RFT. 
```{r}
#| label: fig-bargraph 
#| fig-cap: Bar graph 
#| echo: false
#| warning: false
#| message: false

# Summarize awarded amounts by RFx Type
rfx_summary <- Contract_data %>%
  group_by(RFx_Type) %>%
  summarise(Total_Awarded = sum(Awarded_Amount, na.rm = TRUE)) %>%
  arrange(desc(Total_Awarded))

# Bar plot: Total Awarded Amount by RFx Type
ggplot(rfx_summary, aes(x = RFx_Type, y = Total_Awarded, fill = RFx_Type)) +
  geom_bar(stat = "identity", show.legend = FALSE) +
  scale_y_continuous(labels = dollar_format()) +
  labs(title = "Total Awarded Amount by RFx Type",
       x = "RFx Type",
       y = "Total Awarded Amount ($)") +
  theme_minimal()
```


The pie chart @fig-PieChart demonstrates a significant disparity, with the majority of contracts awarded to large businesses (marked as FALSE for small business). This indicates that small businesses have minimal representation in the awarding process, suggesting a lack of equity.
```{r}
#| label: fig-PieChart
#| fig-cap: Proportion of Small Business Awards
#| echo: false
#| warning: false
#| message: false

# Pie chart: Proportion of Small Business Awards
small_business_summary <- Contract_data %>%
  group_by(Small_Business) %>%
  summarise(Total_Awarded = sum(Awarded_Amount, na.rm = TRUE))

ggplot(small_business_summary, aes(x = "", y = Total_Awarded, fill = Small_Business)) +
  geom_bar(stat = "identity", width = 1, color = "white") +
  coord_polar(theta = "y") +
  labs(title = "Proportion of Small Business Awards",
       x = NULL,
       y = NULL) +
  theme_void()

```



Looking at @fig-Boxplot, categories like "Construction Services" and "Professional Services" show a wide range of awarded amounts, with some outliers receiving significantly higher awards. Smaller categories (e.g., "Offer to Purchase") have fewer instances and seem to award comparatively smaller amounts. This indicates potential concentration of funding in certain categories, which may not benefit small businesses.
```{r}
#| label: fig-Boxplot
#| fig-cap: Awarded Amounts by High-Level Category
#| echo: false
#| warning: false
#| message: false

# Box plot: Awarded Amounts by High-Level Category
ggplot(Contract_data, aes(x = High_Level_Category, y = Awarded_Amount, fill = High_Level_Category)) +
  geom_boxplot(show.legend = FALSE) +
  scale_y_continuous(labels = dollar_format()) +
  labs(title = "Distribution of Awarded Amounts by Category",
       x = "Category",
       y = "Awarded Amount ($)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```


Not working still need to fix 
```{r}
#| label: fig-Time series
#| fig-cap: Awarded Amounts Over Time
#| echo: false
#| warning: false
#| message: false

# Time series: Awarded Amounts Over Time
ggplot(Contract_data, aes(x = Award_Date, y = Awarded_Amount, color = Division)) +
  geom_line() +
  geom_point() +
  scale_y_continuous(labels = dollar_format()) +
  labs(title = "Awarded Amounts Over Time",
       x = "Award Date",
       y = "Awarded Amount ($)") +
  theme_minimal()

```

@fig-bar1, Across most categories, the representation of small businesses is minimal compared to large businesses.Notably, in the "Offer to Purchase" category, there seems to be equal representation of small and large businesses. This could imply specific policies or practices in that category that favor small businesses.

```{r}
#| label: fig-bar1
#| fig-cap: Awarded Amounts Over Time
#| echo: false
#| warning: false
#| message: false


# Calculate proportions of Small and Large Businesses by High-Level Category
category_small_large <- Contract_data %>%
  group_by(High_Level_Category, Small_Business) %>%
  summarise(Total_Awarded = sum(Awarded_Amount, na.rm = TRUE)) %>%
  mutate(Proportion = Total_Awarded / sum(Total_Awarded)) %>%
  ungroup()

# Bar Plot: Proportion of Small vs. Large Businesses by High-Level Category
ggplot(category_small_large, aes(x = High_Level_Category, y = Proportion, fill = Small_Business)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(title = "Proportion of Small vs. Large Businesses by High-Level Category",
       x = "High-Level Category",
       y = "Proportion",
       fill = "Small Business") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))



```


@fig-bar2, we can observe that 
```{r}
#| label: fig-bar2
#| fig-cap: Awarded Amounts Over Time
#| echo: false
#| warning: false
#| message: false

# Calculate proportions of Small and Large Businesses by RFx Type
rfx_small_large <- Contract_data %>%
  group_by(RFx_Type, Small_Business) %>%
  summarise(Total_Awarded = sum(Awarded_Amount, na.rm = TRUE)) %>%
  mutate(Proportion = Total_Awarded / sum(Total_Awarded)) %>%
  ungroup()

# Bar Plot: Proportion of Small vs. Large Businesses by RFx Type
ggplot(rfx_small_large, aes(x = RFx_Type, y = Proportion, fill = Small_Business)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(title = "Proportion of Small vs. Large Businesses by RFx Type",
       x = "RFx Type",
       y = "Proportion",
       fill = "Small Business") +
  theme_minimal()

```




# Discussion

General Inferences:
Equity Gaps: There is a clear lack of equity in the distribution of contracts between small and large businesses. Most of the contracts and larger awards seem to favor larger corporations.
Category-Specific Opportunities: Some categories like "Offer to Purchase" might be worth further investigation to understand what policies support small business equity and whether they can be applied to other categories.
Implications for Policy: These findings could motivate policy adjustments to enhance opportunities for small businesses across all categories, potentially leading to more diverse supplier representation.


## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this. 

## Second discussion point

Please don't use these as sub-heading labels - change them to be what your point actually is.

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

\appendix

# Appendix {-}


# Additional data details

# Model details {#sec-model-details}

## Posterior predictive check

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...

In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows... 

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]

pp_check(first_model) +
  theme_classic() +
  theme(legend.position = "bottom")

posterior_vs_prior(first_model) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom") +
  coord_flip()
```

## Diagnostics

@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...

@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 2

plot(first_model, "trace")

plot(first_model, "rhat")
```



\newpage


# References


